---
output:
  html_document:
    keep_md: yes
  pdf_document:
    keep_tex: yes
---
```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 12, fig.height = 8, fig.path = 'figure/',
                      echo = TRUE, warning = FALSE, message = FALSE)
```
#Statistical Inference
###Course Project 1: CP1_template


###Introduction
This document presents the results of the Course Project for the Coursera course: Statistical Inference. This assessment makes use of simulation techniques in order to explore inference and do some simple inferential data analysis.

The student is to investigate the exponential distribution in R and to draw comparison against the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda.

The student is to investigate the distribution of averages of 40 exponentials, with a thousand simulations.


###1. Load Packages
```{r}
for (package in c('ggplot2')) {
 
    if (!require(package, character.only = TRUE, quietly = FALSE)) {
        install.packages(package)
        library(package, character.only = TRUE)
    }
}
```


###2. Simulations
Set lambda for exponential function, number of exponentials and numbers of sample/tests.
```{r}
set.seed(123456789)

val_lambda <- 0.2
val_n <- 40
val_sims <- 1000
```

Run sample/tests.
```{r}
data_expdist <- data.frame(means = 1:val_sims)

for (i in 1:val_sims) {
 
  val_sim <- rexp(val_n, val_lambda)
  data_expdist[i, 1] <- mean(val_sim)
 
}

remove(val_sim)
```

Find range and plot sampled data.

```{r, echo = FALSE}
val_min <- min(data_expdist$means)
val_max <- max(data_expdist$means)

ggplot(data = data_expdist, aes(x = means)) +
  geom_histogram(binwidth = 0.1, col = "black", alpha = 0.5) +  
  scale_x_continuous(breaks = round(seq(val_min, val_max, by = 1)))
```


###3. Sample Mean vs. Theoretical Mean
Find the theoretical mean of the exponential distribution
```{r}
val_truemu <- 1/val_lambda
val_truemu
```

Find the mean of the sampled distribution
```{r}
val_samplemu <- mean(data_expdist$means)
val_samplemu
```


###4. Sample Variance vs. Theoretical Variance
Find the theoretical variance of the exponential distribution
```{r}
val_truevar <- (1/val_lambda/sqrt(val_n))^2
val_truevar
```

Find the variance of the sampled distribution
```{r}
val_samplevar <- var(data_expdist$means)
val_samplevar
```


###5. Distribution
Plot sampled means against true mean distribution

```{r, echo = FALSE}
ggplot(data = data_expdist, aes(x = means)) +
  geom_histogram(binwidth = 0.1, col = "black", aes(y = ..density..), alpha = 0.5) +
  geom_density(colour = "purple") +
  stat_function(fun = dnorm, arg = list(mean = val_truemu , sd = sqrt(val_truevar)), colour = "blue") +
  geom_vline(xintercept = val_truemu, colour = "purple") +
  geom_vline(xintercept = val_samplemu, colour = "blue") +
  scale_x_continuous(breaks = round(seq(val_min, val_max, by = 1)))
```

Evaluate the coverage of confidence intervals for a standard normal distribution:
```{r}
  mean(data_expdist$means) + c(-1,1) * 1.96 * sd(data_expdist$means)/sqrt(nrow(data_expdist))
```